\chapter{Table of prime numbers to 2000}
\label{app:primes}

In additon to the numbers {2} and {5}, the tables below show all of the prime numbers up to 2000.

\newcounter{hh}
\begin{center}
\begin{tikzpicture}[scale=0.75]
	%% Lefthand headers
	\draw (0,0) grid (1,12);
	\foreach \y in {0,...,9} \draw (0.5, 9.5-\y) node{\large\y};
	\draw (0,11) -- (1,10);
	\draw (0.25,10.25) node{\footnotesize T};
	\draw (0.75,10.75) node{\footnotesize U};
	\draw (0.5,11.5) node{H};
	%% Prime column setup and headers
	\foreach \x in {1,6,11,16}
	{
		%% Units headers
		\draw (\x,0) grid (\x+4,11);
		\draw (\x+0.5,10.5) node{\large1};
		\draw (\x+1.5,10.5) node{\large3};
		\draw (\x+2.5,10.5) node{\large7};
		\draw (\x+3.5,10.5) node{\large9};
		%% Hundreds header
		\draw (\x,11) rectangle (\x+4,12);
		\setcounter{hh}{\x-1};
		\draw (\x+1.4,11.75) node{\arabic{hh}};
		\stepcounter{hh};
		\draw (\x+2.6,11.75) node{\arabic{hh}};
		\stepcounter{hh};
		\draw (\x+2,11.5) node{\arabic{hh}};
		\stepcounter{hh};
		\draw (\x+1.4,11.25) node{\arabic{hh}};
		\stepcounter{hh};
		\draw (\x+2.6,11.25) node{\arabic{hh}};
	}
	%% Primes
	\draw plot[only marks, mark size = 3, mark=*] file {data/primes.txt};
\end{tikzpicture}
\end{center}

\subsection*{How to use the table}

Suppose we want to know whether the number 1439 is prime. Here's how we use the table to determine this:
\begin{enumerate}
\item The box on top of each table shows a group of hundreds (that's what the H stands for at the left side). We should think of the number 1439 as having ``14'' in the hundreds place. The table heading of the third table includes the number 14, so 1439 will appear in this section of the table.

\item The rows of the table indicate the digit in the tens place (that's what the T stands for at the left side). Since 1439 has a 3 in the tens place, we'll use that row of the table.

\item The columns in each table indicate the units place (that's what the U stands for at the left side). We need the column with heading 9, since that's the number in the units place of 1439.

\item We have narrowed our selection down to a single cell in the table. That cell looks like this:
\begin{center}\begin{tikzpicture}
\draw (0,0) rectangle (1,1);
\draw plot[only marks, mark size = 3, mark=*] coordinates {(0.25,0.75)(0.75,0.25)};
\end{tikzpicture}\end{center}
The dots indicate which of the numbers associated with this cell are prime, using the same layout as in the topmost box. This pattern indicates that 1439 is prime. The box also tells us that 1039 is prime, and that 1139, 1239, and 1339 are all composite.
\end{enumerate}

\subsection*{Notice and wonder}

Notice that the columns include only the headings 1, 3, 7, and 9. Why do you suppose that is?

Notice that none of the boxes have all five dots. Why do you suppose that is? If we continue the table, do you suppose we will ever find a box with all five dots? Why or why not?

Notice that if a box contains four dots, it's always the four corners. Why do you suppose that is? If we continue the table, do you suppose we will ever find a box with four dots in a different arrangement? Why or why not?

%%% End of file